Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory

### 2024-07-10 POPL/paper.tex
a better(?) abstract

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Grammar Theoretic Type Theory\xspace}
\newcommand{\theoryabbv}{GTTT\xspace}

@@

\begin{abstract}
  In this paper we propose \theoryname~(\theoryabbv), a new syntactic formalism for formal grammars: a
  dependent linear/non-linear type theory in which the tensor product is
  non-commutative. The linear types in this calculus may be interpreted as
  grammars, and linear terms as parse transformers. This provides a logical
  characterization of grammar classes based on a substructural logic rather than
  ordinary first-order logic.

  Internal to \theoryname, we carry out many classical constructions from
  the theory of formal grammars. Inspired by the common
  structure underlying various ideas in formal language
  theory, our type theory makes these relationships explicit
  and provides a unifying syntax to reason about formal
  grammars.

  We give a semantics in the category of grammars and prove a
  canonicity theorem showing that every term in a context corresponding to a
  fixed string is equal to a term in a canonical form encoding a parse tree of
  that string. Finally, we discuss the application to the design and
  implementation of correct-by-construction parsers.
\end{abstract}

>>>>

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Grammar Type Theory\xspace}
\newcommand{\theoryabbv}{GramTT\xspace}

@@

\begin{abstract}
  We propose \theoryname~(\theoryabbv) a new type-theoretic calculus
  for the specification of formal grammars and intrinsic verification
  of parsers. \theoryabbv is a dependent non-commutative
  linear/non-linear type theory where the linear types can be viewed
  as formal grammars, and linear terms can be viewed as parse
  transformers, of which parsers are a special case. We formalize this
  interpretation by giving a denotational semantics of the theory in a
  novel categorical semantics of formal grammars in a category of
  families indexed by strings.

  To demonstrate the usefulness of the calculus for formal grammar
  theory, we show that different grammar formalisms \max{which ones do
    we actually show?} can be expressed as linear types in
  \theoryabbv. We give simple characterizations of the regular and
  context-free grammars as expressible by restricting which
  connectives of \theoryabbv are allowed, a form of type theoretic
  characterization of these grammar classes. Further, we demonstrate
  some central proofs of formal grammar theory can be intrinsically
  verified: \emph{strong} equivalence between regular expressions and
  NFAs is expressed by a syntactic \emph{isomorphism} definable in
  \theoryabbv.

  Finally, since \theoryabbv is a syntactic calculus, in addition to
  the ``standard'' semantics of linear types as grammars, we
  demonstrate two non-standard semantic interpretations. One is a
  semantics where linear types are instead interpreted as
  \emph{semantic actions}, rather than grammars, which are more common
  in application than formal grammars proper. Lastly we use a
  non-standard semantics to prove a \emph{canonicity} metatheorem that
  shows that the interpretation of types as grammars is canonical: the
  type-theoretic notion of a parse agrees with the expected semantic
  notion.
\end{abstract}
